---
title: "Microbial Competition and Tree Functional Plasticity (2012-2018)"
subtitle: "Long-term Replication and Extension of Jiang et al. (Nature, 2024)"
author: 
  - name: "Analysis Team"
    affiliation: "EucFACE Experiment Analysis"
date: last-modified
date-format: long
format: 
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: show
    embed-resources: true
    df_print: paged
execute:
  warning: false
  message: false
---

# 1. Introduction

<<<<<<< HEAD
## Scientific Context
=======

## 1.1 Scientific Context: The Paradox of Enrichment
<<<<<<< HEAD
>>>>>>> 4826ff6d91e989d1d22185efb273454dd780e59f
=======

Elevated atmospheric CO2 (eCO2) is widely predicted to act as a "fertilizer" for global forests, stimulating photosynthesis and increasing biomass carbon sequestration. This process is relied upon as a critical buffer against climate change in Earth System models. However, empirical evidence suggests that in phosphorus-limited landscapes- which cover vast areas of tropics and subtropics-this fertilization effect often fails to materialize. 

This creates a **"Stoichiometric Mismatch"**: while $eCO_2$ increases the supply of Carbon (C), trees require Phosphorus (P) to convert that carbon into biomass (wood and roots). In ancient, weathered soils, P availability is not governed by simple chemical weathering, but by intense biological competition.  

In the mature Eucalyptus forest of the EucFACE experiment, biomass growth has remained stagnant. Jiang et al. (2024) propose the "Microbial Pre-emption Hypothesis": soil microbes aggressively compete for scarce phosphorus (P), creating a nutrient bottleneck that prevents trees from utilizing the extra carbon.



This analysis replicates and extends Jiang et al. (2024) by shifting the temporal resolution from **annual averages** to **monthly time-steps**. This approach allows us to: 1. Capture rapid microbial pulses masked in annual means. 2. Investigate seasonal "lockout" dynamics. 3. Test for functional plasticity (PUE) with higher statistical power.

Elevated atmospheric CO₂ (eCO₂) typically stimulates plant growth. However, in the mature Eucalyptus forest of the EucFACE experiment, biomass growth has remained stagnant. Jiang et al. (2024) propose the **"Microbial Pre-emption Hypothesis"**: soil microbes aggressively compete for scarce phosphorus (P), creating a nutrient bottleneck that prevents trees from utilizing the extra carbon.

We replicate and extend **Jiang et al. (2024)** using the full available dataset (**2012–2018**). By analyzing this extended 7-year period, we test if the "Lockout" mechanism persists over the long term and if trees exhibit sustained functional plasticity.

## Objectives

1.  **Long-term Trajectories (2012-2018):** Visualize ecosystem dynamics over 7 years to confirm the persistence of biomass stagnation.
2.  **The Lockout:** Validate if Microbial P pools consistently dominate Available Soil P pools over the entire experimental duration.
3.  **Biological Driver:** Test if **P Mineralization Rates** are coupled with microbial biomass.
4.  **Adaptation:** Assess if **Phosphorus Use Efficiency (PUE)** trends shift over the long term.

# 2. Materials & Methods

**Data Origin:**
Data was sourced from the EucFACE repository. We analyze the complete available timeline from **2012 to 2018**, utilizing all overlapping data points for calculations.

**Data Structure:**
1.  **Pools:** Soil Phosphate, Microbial Phosphorus, Wood Carbon, Fine Root Biomass.
2.  **Fluxes:** GPP, P Mineralization, Total Litter Phosphorus Return.
    * *Note: Frass is excluded to prevent temporal bias, as data was unavailable after 2014.*

**Statistical Analysis:**
We employ **Seasonal Aggregation** to synchronize datasets.
* **Time-series:** All available data points are plotted.
* **Correlations & Ratios (PUE):** Calculated only for seasons where paired data exists (Pairwise Deletion).
* **Models:** Linear Mixed-Effects Models (LMM) test for treatment effects: `Variable ~ Treatment * Year + (1|Ring)`.

```{r setup}
#| label: setup-libraries
#| include: false

# 0. CLEAN ENVIRONMENT
rm(list=ls()) 

# 1. LOAD LIBRARIES
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, lme4, lmerTest, knitr, kableExtra, here, patchwork, ggpubr, Hmisc, zoo)

# 2. THEME SETTINGS
theme_set(theme_classic(base_size = 12) + theme(strip.background = element_rect(fill="#f0f0f0")))
cols <- c("Ambient" = "#377EB8", "eCO2" = "#E41A1C")

# 3. HELPER FUNCTIONS
get_trt <- function(ring) { ifelse(ring %in% c(1,4,5), "eCO2", "Ambient") }

get_season <- function(date) {
  m <- month(date)
  case_when(
    m %in% c(12, 1, 2) ~ "Summer",
    m %in% c(3, 4, 5)  ~ "Autumn",
    m %in% c(6, 7, 8)  ~ "Winter",
    TRUE               ~ "Spring"
  )
}

# --- ROBUST MODELING WRAPPER ---
run_safe_lmm <- function(formula, data) {
  if(length(unique(data$Trt)) < 2) return(NULL)
  if(nrow(data) < 4) return(NULL)
  tryCatch({
    lmer(formula, data = data)
  }, error = function(e) { return(NULL) })
}

print_stats <- function(model, caption) {
  if(is.null(model)) return(kable(data.frame(Result="Insufficient Data"), caption=caption))
  anova(model) %>% as.data.frame() %>%
    mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
    mutate(Sig = ifelse(`Pr(>F)` < 0.05, "*", "ns")) %>%
    kable(caption = caption) %>% 
    kable_styling(full_width = F, bootstrap_options = "striped")
}
```


We standardize datasets into a unified Seasonal Time-Series spanning 2012-2018.
```{r}
#| label: data-processing

# --- 1. STRICT LOADER FUNCTION ---
load_file_exact <- function(filename, var_label, type="pool") {
  
  fpath <- list.files(pattern = paste0("^", filename, "$"), recursive = TRUE, full.names = TRUE)[1]
  if(is.na(fpath)) { warning(paste("CRITICAL: File not found:", filename)); return(NULL) }
  
  df <- read_csv(fpath, show_col_types = FALSE)
  
  # Robust Date Parsing
  df <- df %>%
    mutate(Date_Parsed = parse_date_time(as.character(Date), orders = c("ymd", "dmy", "mdy"))) %>%
    filter(!is.na(Date_Parsed))
  
  val_col <- names(df)[grepl("wood|phos|micro|gpp|flux|mg|val|root|min", names(df), ignore.case = T)][1]
  
  df_clean <- df %>%
    rename(RawVal = all_of(val_col)) %>%
    mutate(
      Year = year(Date_Parsed),
      Season = get_season(Date_Parsed),
      Trt = get_trt(Ring),
      Variable = var_label
    )
  
  df_agg <- df_clean %>%
    group_by(Year, Season, Ring, Trt, Variable) %>%
    summarise(
      MeanVal = if(type=="flux") sum(RawVal, na.rm=T) else mean(RawVal, na.rm=T),
      .groups = "drop"
    ) %>%
    mutate(
      SeasonNum = case_when(Season=="Summer"~1, Season=="Autumn"~2, Season=="Winter"~3, Season=="Spring"~4),
      TimeIndex = Year + (SeasonNum-1)/4
    )
  
  return(df_agg)
}

# --- 2. LOAD SPECIFIC FILES ---

# Pools
df_wood   <- load_file_exact("wood_c_pool.csv", "Wood Biomass")
df_soil   <- load_file_exact("soil_phosphate_pool.csv", "Available Soil P")
df_micro  <- load_file_exact("microbial_p_pool.csv", "Microbial P")
df_root   <- load_file_exact("fineroot_c_pool.csv", "Fine Root Biomass")

# Fluxes
df_gpp    <- load_file_exact("overstorey_gpp_flux.csv", "GPP", type="flux")
df_mineral<- load_file_exact("soil_p_mineralization_flux.csv", "P Mineralization", type="flux")

# Litterfall (4 Components)
litter_files <- c("leaflitter_p_flux.csv", "twiglitter_p_flux.csv", "bark_litter_p_flux.csv", "seed_litter_p_flux.csv")
litter_list <- list()

for(fname in litter_files) {
  fpath <- list.files(pattern = paste0("^", fname, "$"), recursive = TRUE, full.names = TRUE)[1]
  if(!is.na(fpath)) {
    d <- read_csv(fpath, show_col_types=F)
    d$Date_Parsed <- parse_date_time(as.character(d$Date), orders = c("ymd", "dmy"))
    d <- d %>% filter(!is.na(Date_Parsed))
    v <- names(d)[grepl("flux|mg", names(d), ignore.case=T)][1]
    litter_list[[fname]] <- d %>% rename(Val = all_of(v)) %>% select(Date_Parsed, Ring, Val)
  }
}

df_litter <- bind_rows(litter_list) %>%
  mutate(Season = get_season(Date_Parsed), Year = year(Date_Parsed)) %>%
  group_by(Year, Season, Ring) %>%
  summarise(TotalP = sum(Val, na.rm=T), .groups="drop") %>%
  mutate(Trt = get_trt(Ring), Variable = "Total P Return") %>%
  rename(MeanVal = TotalP) %>%
  mutate(SeasonNum = case_when(Season=="Summer"~1, Season=="Autumn"~2, Season=="Winter"~3, Season=="Spring"~4),
         TimeIndex = Year + (SeasonNum-1)/4)

# --- 3. MERGE & EXPANDED FILTER (2012-2018) ---
df_final <- bind_rows(df_wood, df_soil, df_micro, df_root, df_gpp, df_mineral, df_litter) %>%
  filter(Year >= 2012 & Year <= 2018) # EXTENDED TIMELINE

kable(table(df_final$Variable, df_final$Year), caption = "Table 1. Data Availability (2012-2018)")
```


# Results & Discussion

##Ecosystem Dynamics (Time-Series Overview)
We first provide a comprehensive view of how all 7 key ecosystem parameters evolved from 2012 to 2016.
```{r}
#| label: fig-overview
#| fig-cap: "Figure 1. Ecosystem Trajectories (2012-2016). Time-series of all key Carbon and Phosphorus variables. Note the distinct seasonal pulses in GPP, Mineralization, and Litter."

# We plot all variables in one large facet grid
ggplot(df_final, aes(x = TimeIndex, y = MeanVal, color = Trt)) +
  stat_summary(fun = mean, geom = "line", size = 0.8) +
  stat_summary(fun = mean, geom = "point", size = 1) +
  stat_summary(fun.data = mean_se, geom = "ribbon", aes(fill = Trt), alpha = 0.1, color = NA) +
  facet_wrap(~Variable, scales = "free_y", ncol = 2) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  labs(x = "Year", y = "Value (Units vary per panel)", title = "Comprehensive Ecosystem Dynamics") +
  theme(legend.position = "top", strip.text = element_text(face="bold"))
```

*Observation:* @fig-overview reveals that while fluxes like GPP and P Mineralization are highly seasonal (peaking in summer), pools like Wood Biomass are relatively stable. Crucially, Available Soil P shows a declining trend over time, coincident with high Microbial P levels, hinting at the lockout mechanism.


## Biomass Stagnation
Has the forest responded to eCO₂?
We verify the primary outcome: Stagnation of wood biomass accumulation.
```{r}
#| label: fig-biomass
#| fig-cap: "Figure 2. Wood Biomass Stagnation. No significant divergence between Ambient and eCO2 treatments."

df_bio <- df_final %>% filter(Variable == "Wood Biomass")
ggplot(df_bio, aes(x = TimeIndex, y = MeanVal, color = Trt)) +
  geom_line(aes(group = interaction(Trt, Ring)), alpha = 0.2) + 
  stat_summary(fun = mean, geom = "line", size = 1.2) +         
  scale_color_manual(values = cols) +
  labs(x = "Year", y = "Wood C (g/m²)", title = "Wood Biomass")

m_wood <- lmer(MeanVal ~ Trt * Year + (1|Ring), data = df_bio)
print_stats(m_wood, "Wood Biomass LMM")
```

*Interpretation:* Wood biomass trajectories overlap completely, and the interaction term is non-significant. Stagnation is confirmed.

## The "Lockout" Mechanism
We test the trade-off between Microbial P and Soil P. Does the microbial pool dominate the soil pool?

```{r}
#| label: fig-lockout
#| fig-cap: "Figure 3. The Lockout. (A) Microbial P dominates Soil P. (B) Negative correlation confirms drawdown."

df_mech <- df_final %>% filter(Variable %in% c("Microbial P", "Available Soil P"))
df_wide <- df_mech %>% pivot_wider(names_from = Variable, values_from = MeanVal) %>% na.omit()

p1 <- ggplot(df_mech, aes(x = TimeIndex, y = MeanVal, color = Trt)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  facet_wrap(~Variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = cols) +
  theme(legend.position = "none")

p2 <- ggplot(df_wide, aes(x = `Microbial P`, y = `Available Soil P`, color = Trt)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(method = "pearson") +
  scale_color_manual(values = cols)

p1 + p2
```

## Biological Driver: P Mineralization
*Extension:* We analyze the Biological Driver (Mineralization) to see if microbial biomass drives nutrient turnover.

```{r}
#| label: fig-mineral
#| fig-cap: "Figure 4. P Mineralization Dynamics. (A) Strong seasonal pattern with summer peaks. (B) Positive correlation with Microbial P confirms biological drivers."

# A. Seasonal Pattern (Boxplot)
df_min <- df_final %>% filter(Variable == "P Mineralization")

p3 <- ggplot(df_min, aes(x = Season, y = MeanVal, fill = Trt)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = cols) +
  labs(title = "A. Seasonal Mineralization", y = "Flux Rate")

# B. Correlation (Microbes vs Mineralization)
df_bio_link <- df_final %>% 
  filter(Variable %in% c("Microbial P", "P Mineralization")) %>%
  pivot_wider(names_from = Variable, values_from = MeanVal) %>%
  na.omit()

p4 <- ggplot(df_bio_link, aes(x = `Microbial P`, y = `P Mineralization`, color = Trt)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  stat_cor(method = "pearson") +
  scale_color_manual(values = cols) +
  labs(y = "P Mineralization Flux", x = "Microbial P", title = "B. Biological Control")

p3 + p4
```

## Functional Adaptation (PUE)
We examine Adaptation via PUE trends over Time and Season.

```{r}
#| label: fig-pue
#| fig-cap: "Figure 5. Phosphorus Use Efficiency (PUE). (A) Long-term trend shows increasing efficiency under eCO2. (B) Seasonal analysis shows adaptation is strongest in active seasons."

df_pue <- df_final %>%
  filter(Variable %in% c("GPP", "Total P Return")) %>%
  pivot_wider(names_from = Variable, values_from = MeanVal) %>%
  mutate(PUE = GPP / `Total P Return`) %>%
  filter(is.finite(PUE) & PUE > 0 & PUE < 6000)

# A. Time Series
p5 <- ggplot(df_pue, aes(x = TimeIndex, y = PUE, color = Trt)) +
  stat_summary(fun = mean, geom = "line", size = 1) +
  stat_summary(fun = mean, geom = "point", size = 1.5) +
  scale_color_manual(values = cols) +
  labs(title = "A. PUE Trends (2012-2016)", x = "Year") +
  theme(legend.position = "none")

# B. Seasonal Boxplot
p6 <- ggplot(df_pue, aes(x = Season, y = PUE, fill = Trt)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = cols) +
  labs(title = "B. Seasonal Adaptation", x = NULL)

p5 / p6
```

```{r}
#| label: tbl-pue
#| tbl-cap: "Table 3. LMM Results for PUE."
m_pue <- lmer(PUE ~ Trt * Season + (1|Ring), data = df_pue)
print_stats(m_pue, "PUE LMM Results")
```

# Conclusion
This expanded analysis (2012-2016) confirms the Microbial Pre-emption Hypothesis:

*Overview:* Whole-ecosystem monitoring reveals consistent nutrient limitation patterns (@fig-overview).

*Mechanism:* Microbes dominate the P pool (@fig-lockout) and drive mineralization rates (@fig-mineral).

*Adaptation:* Trees under eCO2 exhibit clear functional plasticity (higher PUE), particularly in growing seasons (@fig-pue), yet this is insufficient to break the biomass stagnation.