---
title: "Microbial Competition and Tree Functional Plasticity"
subtitle: "Extended Analysis: Monthly Dynamics & Seasonality (2013-2021)"
author: 
  - name: "Gia Trieu Pham; Ani Ugrelidze; Oishanu Dutta; Muhammad Hasher Naveed"
    affiliation: Universitat Politècnica de València
date: last-modified
date-format: iso
format: 
  html:
    theme: cosmo
    toc: true
    number-sections: true
    code-fold: true
    embed-resources: true
    df_print: paged
execute:
  warning: false
  message: false
---

# Introduction & Setup

This analysis replicates and extends Jiang et al. (2024) by shifting the temporal resolution from **annual averages** to **monthly time-steps**. This approach allows us to: 1. Capture rapid microbial pulses masked in annual means. 2. Investigate seasonal "lockout" dynamics. 3. Test for functional plasticity (PUE) with higher statistical power.

We employ **Linear Interpolation** to align asynchronous datasets (e.g., annual biomass measurements vs. monthly litterfall) for continuous multivariate analysis.

```{r}
#| label: setup-libraries
#| include: false

# 0. CLEAN ENVIRONMENT (Fixes 'environment' errors)
rm(list=ls()) 

# 1. LOAD LIBRARIES
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, lubridate, lme4, lmerTest, knitr, kableExtra, here, zoo, patchwork)

# 2. VISUALIZATION THEME
theme_set(theme_classic(base_size = 12))
cols <- c("Ambient" = "#377EB8", "eCO2" = "#E41A1C")

# 3. HELPER FUNCTION FOR ANOVA TABLES
run_anova <- function(model) {
  anova(model) %>% 
    as.data.frame() %>%
    mutate(across(where(is.numeric), \(x) round(x, 3))) %>%
    mutate(Sig = ifelse(`Pr(>F)` < 0.05, "*", "")) %>%
    kable() %>% kable_styling(full_width = F)
}
```

# Data Processing: Monthly Aggregation & Interpolation

We aggregated data to a monthly time-step to capture seasonal dynamics. Linear interpolation (`zoo::na.approx`) was used to fill sampling gaps, ensuring continuous time-series for multivariate analysis.

```{r}
#| label: load-clean-data

# --- STEP 1: SIMPLIFIED DATA LOADING FUNCTION ---
get_monthly_data <- function(file, pattern, is_flux = FALSE) {
  # Locate file (checks multiple directories)
  path <- here("EucFACE_P_Analysis/data", file)
  if(!file.exists(path)) path <- here("data", file)
  
  df <- read_csv(path, show_col_types = FALSE)
  
  # Identify value column based on pattern
  val_col <- names(df)[grepl(pattern, names(df), ignore.case = TRUE)][1]
  
  df %>%
    mutate(Date = parse_date_time(Date, orders = c("ymd", "dmy", "mdy"))) %>%
    filter(!is.na(Date)) %>%
    # CRITICAL: Floor date to the 1st of the month for synchronization
    mutate(MonthDate = floor_date(Date, "month")) %>% 
    rename(RawValue = all_of(val_col)) %>%
    # Group by Month and Ring
    group_by(MonthDate, Ring) %>%
    summarise(Val = if(is_flux) sum(RawValue, na.rm=T) else mean(RawValue, na.rm=T), .groups="drop")
}

# --- STEP 2: LOAD INDIVIDUAL VARIABLES ---
# Pools (Inventory) -> Use Mean
wood  <- get_monthly_data("wood_c_pool.csv", "wood") %>% rename(Wood_C = Val)
soil  <- get_monthly_data("soil_phosphate_pool.csv", "phosphate") %>% rename(Soil_P = Val)
micro <- get_monthly_data("microbial_p_pool.csv", "microbial") %>% rename(Micro_P = Val)
root  <- get_monthly_data("fineroot_c_pool.csv", "fineroot") %>% rename(Root_C = Val)
gpp   <- get_monthly_data("overstorey_gpp_flux.csv", "GPP") %>% rename(GPP = Val)

# Fluxes (Litterfall) -> Use Sum
litter_files <- c("leaflitter_p_flux.csv", 
  "twig_litter_p_flux.csv", 
  "bark_litter_p_flux.csv", 
  "seed_litter_p_flux.csv") 
litter <- lapply(litter_files, function(f) get_monthly_data(f, "flux", is_flux = TRUE)) %>%
  bind_rows() %>%
  group_by(MonthDate, Ring) %>%
  summarise(Litter_P = sum(Val, na.rm=T), .groups="drop")

# --- STEP 3: MERGE & INTERPOLATE (CRITICAL) ---
df_final <- list(wood, soil, micro, root, gpp, litter) %>%
  reduce(full_join, by = c("MonthDate", "Ring")) %>%
  mutate(Ring = factor(Ring)) %>%
  
  # Arrange for correct time-series interpolation
  arrange(Ring, MonthDate) %>%
  
  # Create a complete time grid for every Ring (Fixes missing months)
  group_by(Ring) %>%
  complete(MonthDate = seq(min(MonthDate), max(MonthDate), by = "month")) %>%
  
  # Linear Interpolation (Fill NA values based on neighbors)
  mutate(across(c(Wood_C, Soil_P, Micro_P, Root_C, GPP, Litter_P), 
                ~na.approx(.x, na.rm = FALSE, rule = 2))) %>%
  ungroup() %>%
  
  # Add Metadata & Derived Variables
  mutate(
    Year = year(MonthDate),
    # Define Australian Seasons (Southern Hemisphere)
    Season = case_when(
      month(MonthDate) %in% c(12, 1, 2) ~ "Summer",
      month(MonthDate) %in% c(3, 4, 5) ~ "Autumn",
      month(MonthDate) %in% c(6, 7, 8) ~ "Winter",
      TRUE ~ "Spring"
    ),
    Trt = ifelse(Ring %in% c(1,4,5), "eCO2", "Ambient"),
    PUE = GPP / Litter_P # Phosphorus Use Efficiency
  ) %>%
  filter(Year >= 2012 & Year <= 2018) # Filter for the consistent data period

# Preview Data
head(df_final, 10)
```

# Results: Replication (The Mechanism & Consequence)

## A. The Consequence: Biomass Stagnation

First, we verify the "Stagnation Hypothesis": Does Wood Carbon increase over time under eCO2?

```{r}
#| label: fig-biomass
#| fig-cap: "Biomass Trend. The eCO2 trajectory (Red) does not significantly diverge from Ambient (Blue) over time."

ggplot(df_final, aes(x = MonthDate, y = Wood_C, color = Trt)) +
  # Faint lines for individual rings
  geom_line(aes(group = Ring), alpha = 0.3) +
  # Overall trend line (Linear Model)
  geom_smooth(method = "lm", se = TRUE, size = 1.2) +
  scale_color_manual(values = cols) +
  labs(x = "Time (Month)", y = "Wood Carbon (g/m²)", title = "Wood Biomass Trajectory")

#| label: stats-biomass
# Model: Biomass ~ Treatment * Year
m_wood <- lmer(Wood_C ~ Trt * Year + (1|Ring), data = df_final)
run_anova(m_wood)
```

**Observation:** The overlapping trend lines and the non-significant interaction term in the ANOVA table confirm that eCO2 has not unlocked additional biomass growth.

## B. The Mechanism: Microbial Lockout

Why is biomass stagnating? We test the "Microbial Pre-emption" hypothesis by examining the relationship between Microbial P and Soil P.

```{r}
#| label: fig-lockout
#| fig-cap: "Negative Correlation. Increasing microbial P (x-axis) is associated with reduced available soil P (y-axis)."

ggplot(df_final, aes(x = Micro_P, y = Soil_P, color = Trt)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +
  scale_color_manual(values = cols) +
  labs(x = "Microbial P (g/m²)", y = "Soil Available P (g/m²)", title = "Microbial Drawdown Mechanism")

#| label: stats-lockout
# Model: Soil P ~ Microbial P
m_lockout <- lmer(Soil_P ~ Micro_P * Trt + (1|Ring), data = df_final)
run_anova(m_lockout)
```

**Observation:** The significant negative slope supports the hypothesis that microbes are actively sequestering phosphorus, limiting availability for trees.

## C. Seasonality of Competition

When is the competition most intense? We break down microbial dynamics by season to explore when the competition is most intense and if trees show functional plasticity (PUE).

```{r}
#| label: fig-season-combined
#| fig-cap: "Left: Microbial P pools peak in Summer. Right: Trees tend to increase PUE in Summer/Autumn to adapt to scarcity."

p1 <- ggplot(df_final, aes(x = Season, y = Micro_P, fill = Trt)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = cols) +
  labs(title = "1. Seasonal Microbial P", x = NULL, y = "Microbial P (g/m²)") +
  theme(legend.position = "none")

p2 <- ggplot(df_final %>% filter(PUE < 1000), aes(x = Season, y = PUE, fill = Trt)) +
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values = cols) +
  labs(title = "2. Seasonal PUE (Adaptation)", x = NULL, y = "PUE (C (g) / P (g) )")

p1 + p2 # Combine plots side-by-side

#| label: stats-pue
# Model: PUE ~ Treatment * Season
m_pue <- lmer(PUE ~ Trt * Season + (1|Ring), data = df_final)
run_anova(m_pue)
```

# Conclusion

By extending the analysis to monthly time-steps with linear interpolation, we observe:

1.  **Stagnation:** Wood biomass remains unresponsive to eCO2.

2.  **Lockout:** There is a clear negative relationship between microbial biomass and soil P availability.

3.  **Seasonality:** Microbial competition peaks in Summer. Trees exhibit signs of adaptation (increased PUE) during this period, but it is insufficient to overcome the nutrient limitation.
